% Copyright INRIA

\addcontentsline{toc}{chapter}{\string\numberline\space \ Bibliography}
\begin{thebibliography}{99}

\label{ref}

\bibitem{Abram} M. Abramowitz, I.A. Stegun, eds., {\em Handbook of Mathematical Functions}, National Bureau of Standards Applied Mathematics Series, no. 55, Superintendent of Documents, U.S. Goverment Printing Office, Washington, D.C., 1964.

\bibitem{anderson} B.D.O. Anderson and J.B. Moore, {\em Optimal Filtering}, Englewood Cliffs, NJ: Prenctice Hall, 1979.

\bibitem{Bier} Gerald J. Bierman, {\em Factorization Methods for Discrete Sequential Estimation}, Academic Press, New York, 1977.


\bibitem{Carlson} B.C. Carlson, {\em Special Functions of Applied Mathematics}, Academic Press, New York, 1977.

\bibitem{cheney} E.W. Cheney, {\em Introduction to Approximation Theory}, McGraw-Hill, 1966.

\bibitem{claas} T.A.C.M. Claasen and W.F.G. Mecklenbrauker, ``The Wigner distribution: a tool for time frequency signal analysis'', {\em Phillips J. Res.}, vol. 35, pp. 217-250, 276-300, 372-389, 1980.

\bibitem{delebecque} F. Delebecque, C. Klimann and S. Steer, {\em Basile, guide d'utilisation}, Internal Report INRIA, 1987.

\bibitem{dieu} J. Dieudonn\'e, {\em Calcul Infinit\'esimal}, Hermann, Paris, 1980.

\bibitem{faurre} P. Faurre, {\em R\'ealisations markoviennes de processus stationnaires \em}, Th\`ese d'\'etat, Rapport Laboria n0 13, 1973.

\bibitem{Gelb} A. Gelb, ed., {\em Applied Optimal Estimation}, MIT Press, 1974.

\bibitem{HB} Francis B. Hildebrand, {\em Methods of Applied Mathematics}, Prentice Hall, Englewood Cliffs, N.J., 1965.

\bibitem{ho} B.L. Ho and R.E. Kalman, ``Effective construction of linear state variable models from Input/Output data'', {\em Proc. 3rd Allerton Conference}, 1963.

\bibitem{handj} R. Hooke, T.A. Jeeves, ``Direct Search solution of numerical and statistical problems'', Journ. Assoc. Comput. Mach., Vol 8 No 2, pp 212-229, April 1961.

\bibitem{Kailath1} T. Kailath, {\em Lectures on Wiener and Kalman Filtering}, Springer-Verlag, New York, 1981.

\bibitem{Kailath2} T. Kailath, {\em Linear Systems}, Prentice Hall, 1980.

\bibitem{KM} Steven M. Kay and Stanley L. Marple, ``Spectrum Analysis - A Modern Perspective'', {\em Proc. IEEE}, vol. 69, no. 11, pp. 1380-1419, Nov. 1981.

\bibitem{kung} S. Kung, ``A new identification and model reduction algorithm via singular value decompositions'', Proc. 12th Asilomar Conf. Circuits, Syst. Comput., Pacific Grove, CA, pp 705-714, Nov 1978.


\bibitem{flandrin} W. Martin and P. Flandrin, ``Wigner-Ville spectral analysis of nonstationary processes'', {\em IEEE Trans. Acoust., Speech, Signal Processing}, Vol. ASSP-33, N0 6, Dec. 1985.

\bibitem{matlab}  C. Moler, {\em MATLAB User's  Guide},  Tech.  Rep.  CS81-1, Dept.  of Computer Sci., Univ.  New Mexico, August, 1982.


\bibitem{Nikoukhah} R. Nikoukhah, {\em A Deterministic and Stochastic Theory of Two-Point Boundary-Value Descriptor Systems}, Ph.D. dissertation, Dept. of Elec. Eng. and Comp. Science, M.I.T., 1988.

\bibitem{oppen} A.V. Oppenheim and R.W. Schafer, {\em Digital Signal Processing}, Prentice Hall, 1975.

\bibitem{outils} M. Prevosto, A. Benveniste, B. Barnouin, ``La m\'ethode des variables instrumentales en treillis; application \`a la mod\'elisation des structures en vibration'', Outils et mod\`eles math\'ematiques pour l'Automatique, l'Analyse des syst\`emes et le Traitement du Signal, Vol 2, pp 639-667, Ed. CNRS, 1982.

\bibitem{ieee} {\em Programs for Digital Signal Processing}, IEEE Press John Wiley and Sons, 1979.

\bibitem{rabiner} L. Rabiner and B. Gold, {\em Theory and Application of Digital Signal Processing}, Prentice Hall, 1975.

\bibitem{czt} L.R. Rabiner, R.W. Schafer, and C.M. Rader, ``The Chirp Z-Transform Algorithm and its Applications'', {\em Bell Sys. Tech. J.}, vol 48, pp. 1249-1292, May 1969.

\bibitem{rissanen} J. Rissanen, ``Recursive identification of linear systems'', {\em SIAM Journal of Control}, Vol 9, 1971.

\bibitem{ruckebusch} G. Ruckebusch, {\em Repr\'esentations markoviennes de processus gaussiens stationnaires}, Th\`ese 3\`eme cyle math. stat., Paris VI, 1975.

\bibitem{steiglitz} K. Steiglitz, ``Designing Short-Word recursive digital filters'', Proc. 9th Allerton Conf. on Circuit and Systems Theory, pp 778-788, Oct. 1971.

\bibitem{Stewart} G.W. Stewart, {\em Introduction to Matrix Computations}, Academic Press, New York, 1973.

\bibitem{VanLoan} C.F. Van Loan,  ``Computing integrals involving matrix exponentials'', {\em IEEE Trans. Autom. Control}, vol AC-23, pp. 395-404, 1978.

\end{thebibliography}
