% Copyright INRIA

\chapter{Spectral Estimation}
\section{Estimation of Power Spectra}
\index{spectral estimation}

	The power spectrum of a deterministic,
finite length, discrete-time signal, $x(n)$, is defined to
be the magnitude squared of the signal's Fourier transform
%
\begin{equation}
S_x(\omega)=\frac{1}{N}|\sum_{n=0}^{N-1}x(n)e^{-j\omega n}|^2.
\label{e.sp.1}
\end{equation}
%
In an analogous fashion the cross-spectrum of two signals 
$x(n)$ and $y(n)$ is defined to be
%
\begin{equation}
S_{xy}(\omega)=\frac{1}{N}(\sum_{n=0}^{N-1}x(n)e^{-j\omega n})(\sum_{n=0}^{N-1}y(n)e^{-j\omega n})^*.
\label{e.sp.2}
\end{equation}
%
The power spectra of random, zero-mean, wide sense stationary signals
are obtained from the Fourier transform of the correlation functions
of these signals.  Thus, for $R_x$ representing the autocorrelation
function of $x$ and $R_{xy}$ representing the cross-correlation function of
$x$ with $y$ we have by definition that
%
\begin{eqnarray}
R_x(m)&=&E\{x(n+m)x^*(n)\}\label{e.sp.3}\\
R_{xy}(m)&=&E\{x(n+m)y^*(n)\}.
\label{e.sp.4}
\end{eqnarray}
%
Consequently, the power spectrum and cross-power spectrum 
of $x(n)$ and of $x(n)$ with $y(n)$ are, respectively,
%
\begin{eqnarray}
S_x(\omega)&=&\sum_{m=-\infty}^{\infty}R_x(m)e^{-j\omega m}
\label{e.sp.5}\\
S_{xy}(\omega)&=&\sum_{m=-\infty}^{\infty}R_{xy}(m)e^{-j\omega m}.
\label{e.sp.6}
\end{eqnarray}
%

The formulas (\ref{e.sp.5}) and (\ref{e.sp.6})  require estimates of 
the correlation functions.
Possible candidates for the estimates of the auto and cross
correlation functions of finite length random signals
(i.e., $x(n)\ne0$ and $y(n)\ne0$ for $n=0,1,\ldots,N-1$) are
%
\begin{eqnarray}
\hat{R}_x(m)&=&\frac{1}{N}\sum_{n=0}^{N-1-m}x(n+m)x^*(n)
\label{e.sp.7}\\
\hat{R}_{xy}(m)&=&\frac{1}{N}\sum_{n=0}^{N-1-m}x(n+m)y^*(n).
\label{e.sp.8}
\end{eqnarray}
%
The estimates in (\ref{e.sp.7}) and (\ref{e.sp.8}) are unbiased, 
consistent estimators in the limit as $N\rightarrow\infty$.  
Furthermore, in the 
case where the random signals are jointly Gaussian, these
estimators are the maximum likelihood estimates of the correlation
functions.  Another interesting property of 
the estimators in (\ref{e.sp.7}) and (\ref{e.sp.8}) is that
when substituted, respectively, into the expressions in 
(\ref{e.sp.5}) and (\ref{e.sp.6}), after some algebraic
manipulation, yield exactly the expressions in (\ref{e.sp.1})
and (\ref{e.sp.2}).

	Unfortunately, there is a serious problem with the 
above power spectrum estimation scheme.  This problem is that the 
resulting power spectral estimates, in the limit, are
not consistent.  That is, the error variance
of the estimate does not decrease with increasing data.
Consequently, power spectral estimates obtained
by taking the magnitude squared of the Fourier transform
are high-variance, low-quality estimates.

	In the sections which follow two
techniques are discussed which yield improved
spectral estimates.  These techniques are both based on averaging
spectral estimates obtained
from the classical approach just described.  This
averaging, although introducing some biasing,
yields greatly improved estimates in that, in the
limit, these estimates become consistent.

	The first averaging technique also sections
the data into overlapping segments.  However, in this
case the magnitude squared of the Fourier
transform is calculated from each segment and then these
are averaged together to yield the spectral estimate.
This technique is called the modified periodogram
method for spectral estimation.

	The second averaging technique sections the data
into overlapping segments.  For each segment an estimate of
the correlation function is calculated.  These estimates are
then averaged and the estimated power spectral density
is the Fourier transform of the average.  This technique is known
as the correlation method for spectral estimation.

	Both techniques use windows\index{windowing!spectral estimation}
to diminish the effects that finite data has on spectral estimation.
These effects are identical to the problems encountered in
FIR filter design, and, consequently, the reader is referred to
the FIR filter design section for an explanation of the
issues involved in the choice of windows.  In the discussion
which follows cross-spectral estimation is not discussed considering
that the cross-spectral estimate can be obtained as a simple
modification of the auto-spectral estimation techniques.

\section{The Modified Periodogram Method}
\index{periodogram method}
\index{spectral estimation!periodogram method}

	The modified periodogram method of spectral estimation
repeatedly calculates the periodogram of windowed sub-sections of the data.
These periodograms are then averaged together and normalized by
an appropriate constant to obtain the final spectral estimate.
It is the averaging process which reduces the variance in the
estimate.

	The periodogram of a finite data sequence is defined by
%
\begin{equation}
I(\omega)=\frac{1}{U}|\sum_{n=0}^{N-1}w(n)x(n)e^{-j\omega n}|^2.
\label{e.sp.9}
\end{equation}
%
where $w(n)$ is a window function which has energy $U$.  Consequently,
if $K$ sub-segments of length $N$ are used to calculate the spectrum
of the signal then the modified periodogram spectral estimate, $\hat{S}_x$,
is just the average of the $K$ periodograms
%
\begin{equation}
\hat{S}_x(\omega)=\frac{1}{K}\sum_{k=0}^{K-1}I_k
\label{e.sp.10}
\end{equation}
%
where each of the $I_k$ is the periodogram (calculated as in (\ref{e.sp.9}))
of the $k^{th}$ segment of data.

	Normally, the $K$ segments of data are taken so that there
is a regular pattern of overlap in the successive segments.  That is,
the $k^{th}$ and $(k+1)^{th}$ segments overlap by $D$ points.
Figure~\ref{f.spect1} illustrates three consecutive
overlapping segments of a data sequence.
%
\input{\Figdir spect1.tex}
\dessin{{\tt exec('spect1.code')} Overlapping Data}{f.spect1}
%
It can be shown that an overlap of fifty percent in the data segments
results in an approximately fifty percent reduction in the variance
of the estimate of the power spectrum \cite{rabiner}.  Normally, one
chooses the length of the data segments to reflect the {\em a priori}
knowledge of the correlation length of the data.  That is to say
that if the correlation between two data samples separated by more than $M$
points is considered negligible then the data segment should be of
a length on the order of $M$.  The number of data segments used determines
the variance of the spectral estimate.  The variance decreases proportionally
to the number of independent segments.  Of course, with a limited
quantity of data the number of data segments must also be limited.

	The function {\tt pspect} calculates an estimate of
the power spectrum using the modified periodogram method.

\subsection{Example Using the {\tt pspect} function}
\index{periodogram method!example}

	In this section, we demonstrate the use of the {\tt pspect}
macro.  The data used is generated by passing zero-mean white noise
of unit variance through a low-pass filter.  Consequently, the spectrum
of the data should be the magnitude square of the filter frequency
response.  The low-pass filter is an FIR filter of length 33 generated
using the function {\tt eqfir} .  

	The data was generated using the following Scilab commands,
\verbatok{\Diary pspect1.dia}
\end{verbatim}
As can be seen, a total of 1024 data points are available for the
estimation of the spectrum.  The logarithm of the magnitude squared
of the filter frequency response is shown in Figure~\ref{f.spect2}.
%
\input{\Figdir spect2.tex}
\dessin{{\tt exec('spect2\_4.code')} Log Magnitude Squared of Filter}{f.spect2}
%


	The data obtained above are used to estimate the power spectrum
in the following way
\begin{verbatim}
-->[sm]=pspect(100,200,'tr',y);
\end{verbatim}
The log-magnitude of the power spectrum ({\tt sm}) is plotted in
Figure~\ref{f.spect3}.  It should be pointed 
out here that the value of the section lengths was chosen to be
200 in this example to obtain additional resolution in the 
plot of the Fourier transform of the estimated power spectrum in
Figure~\ref{f.spect3}.  However, there is very acceptable
behavior of the spectral estimate when the section length is
on the order of twice the filter length.  This is due to the
fact that one does not expect correlations in the data that are 
much longer than 
the filter length.  Normally, the section lengths are chosen to
reflect the {\em a priori} knowledge of the correlation length in
the data.
%
\input{\Figdir spect3.tex}
\dessin{{\tt exec('spect2\_4.code')} Estimate of Spectrum}{f.spect3}
%
As can be seen the estimated spectrum matches the theoretical spectrum
(Figure~\ref{f.spect2}) very well.  In particular, 
the peaks of the spectrum in both the pass
and stop bands matches well with those of the filter magnitude response.
Furthermore, the normalization of the estimate is accurate with respect
to the filter response.


\section{The Correlation Method}
\index{spectral estimation!correlation method}
\index{correlation method}

	The correlation method for power spectral estimation
calculates the spectral estimate as the Fourier transform
of a modified estimate of the autocorrelation function.
This modified estimate of the autocorrelation function
consists of repeatedly calculating estimates of the 
autocorrelation function as in (\ref{e.sp.7}) from
overlapping sub-segments of the data, and then averaging
these estimates to obtain the modified estimate.

	Consequently, referring again to Figure~\ref{f.spect1},
for each length $N$ sub-segment of the data $x_k(n)$ the estimate
of $2M$ points of the autocorrelation function is calculated by
%
\begin{equation}
\hat{R}_k(m)=\sum_{n=0}^{N-1-m}x(n+m)x^*(n)
\label{e.sp.11}
\end{equation}
%
for $m=0,\pm1,\pm2,\ldots,\pm M$.  For $K$ estimates of the autocorrelation 
function calculated as in (\ref{e.sp.11}) the power spectral estimate is
obtained from
%
\begin{equation}
\hat{S}_x(\omega)={\cal F}\{\tilde{R}_x(m)w(m)\}
\label{e.sp.12}
\end{equation}
%
where ${\cal F}\{\cdot\}$ represents the Fourier transform operation,
$w(m)$ is a window function, and $\tilde{R}_x(m)$ is the
average of the $K$ estimates
%
\begin{equation}
\tilde{R}_x=\frac{1}{K}\sum_{k=1}^{K}\hat{R}_k.
\label{e.sp.13}
\end{equation}
%

	The correlation method of spectral estimation is based
on the {\tt corr} primitive in Scilab.  The primitive {\tt corr}
is useful for any application requiring correlations or cross-correlations.
Documentation on this primitive can be found in the introductory
manual for Scilab.

The function {\tt cspect} calculates an estimate of
the power spectrum using the correlation method for spectral estimation.

\subsection{Example Using the function {\tt cspect}}
\index{correlation method!example}

	Here, for comparison purposes, the same example as used in
the case of the {\tt pspect} macro is examined using the {\tt cspect}
macro.  The data used is identical
to that used in the previous example.  These data 
are used to estimate the power spectrum
in the following way
\begin{verbatim}
-->[sm]=cspect(100,200,'tr',y);
\end{verbatim}
The log-magnitude of the power spectrum ({\tt sm}) is plotted in
Figure~\ref{f.spect4}.  

It should be pointed 
out here that the value of the 
the number of lags (100) and the number of transform points (200)
were chosen to match the previous example where the {\tt pspect}
macro was used.
A plot of the estimated power spectrum is illustrated in
Figure~\ref{f.spect4}.
%
\input{\Figdir spect4.tex}
\dessin{{\tt exec('spect2\_4.code')} Estimate of Spectrum}{f.spect4}
%
As can be seen the estimated spectrum also matches the theoretical spectrum
(Figure~\ref{f.spect2} very well.  There are some differences,
however, between the estimates obtained using the two different macros.
The primary difference to keep in mind is the difference in how the windows
are used for the two different techniques.  In the correlation
method the magnitude of the window is convolved with the
spectrum of the signal.  In the modified periodogram method the {\em square}
of the magnitude of the window is convolved with the spectrum of the signal.
Consequently, the effects of windows are different in each of the two cases
(for example, the side-lobes of the window are lower in the case of the 
modified periodogram method due to the squaring of its magnitude).
The quantitative differences between the two techniques are difficult
to address here due to the complexity of the question.  There are some
relevant questions concerning which technique may be the best in any
one application.  For more information on how to choose between the
techniques the user is referred to \cite{rabiner} and the relevant references
found within.


\section{The Maximum Entropy Method}
\index{spectral estimation!maximium entropy method}
\index{maximum entropy method}
\subsection{Introduction}

	The power spectrum of a deterministic signal
is defined to be the squared magnitude of the signal's Fourier
transform.  That is, for $x(n)$ a discrete time signal, the power spectrum,
$S_x(\omega)$, is 
%
\begin{equation}
     S_x(\omega)=|X(\omega)|^2
\label{e33.1}
\end{equation}
%
where
%
\begin{equation}
     X(\omega)=\sum_{n=-\infty}^{\infty}x_ne^{-j\omega n}.
\label{e3.2}
\end{equation}
%
In many applications it is very useful to know
the power spectrum of a signal, however, it is
rare that the obtained signal can be characterized as being
deterministic.  Often the signal is present in a noisy
environment and, in addition, is obtained with instrumentation
which degrades the signal with measurement noise.  Consequently, for a 
non-deterministic signal one seeks to estimate the power spectrum.
It can be shown \cite{oppen} that taking
the Fourier transform of a non-deterministic
signal in order to estimate its power spectrum is a very poor
approach.  The problem is that the resulting power spectrum
is a highly variable estimate of the true power spectrum
and that the variance does not decrease to zero as one
increases the data length (i.e., the estimator is not consistent).

	The problem of estimating the power spectrum
can be modified as follows.  Let
$x(n)$ be a zero-mean, stationary signal and let $r_x(n)$ be the autocorrelation
function of the signal (that is, $r_x(n)=E\{x(k)x^*(n+k)\}$).
Then the power spectrum $S_x(n)$ of $x(n)$ is taken to be the Fourier transform
of $r_x(n)$
%
\begin{equation}
     S_x(\omega)=\sum_{n=-\infty}^{\infty}r_x(n)e^{-j\omega n}.
\label{e3.3}
\end{equation}
%
Assuming that statistical averages of the signal $x(n)$ are equal to
time averages we can take as an estimate for $r_x(n)$
%
\begin{equation}
     \hat{r}_x(n)=\lim_{N\rightarrow\infty}\frac{1}{2N+1}\sum_{m=-N}^{N}x(m)x(m-n).
\label{e3.4}
\end{equation}
%
However, after plugging (\ref{e3.4}) into (\ref{e3.3}) and performing some
algebraic manipulation it can be seen that (\ref{e3.3}) is just
the magnitude squared of the Fourier transform of $x(n)$.
Consequently, (\ref{e3.3}) is not any more useful than (\ref{e33.1}) for estimating the
power spectrum of a non-deterministic signal.

	One can improve the statistical properties of 
the power spectrum estimate by smoothing the result obtained
in (\ref{e3.3}) or by breaking the input, $x(n)$, into 
pieces, performing the calculation in (\ref{e3.4}) and (\ref{e3.3}) for
each piece, and then averaging these results together.
These approaches are the classical methods of power spectral
estimation.

	These classical methods of power spectral estimation
are undesirable for two reasons.  The estimate obtained from
these methods is based on a finite (i.e., windowed) segment of the
autocorrelation function.  This means that the resulting estimated power 
spectrum is a convolution of the true power spectrum and of the Fourier transform
of the window.  Consequently, the resolution of spectral features is
diminished and the estimate of the power spectrum at any point
is biased by leakage from other parts of the power spectrum through the 
window sidelobes.

	The maximum entropy spectral estimate (MESE)
of the power spectrum yields an improved spectral
density estimate.  That's to say that for MESE
the resolution is improved and the bias
is decreased in comparison with the classical spectral estimation
methods.  This improvement is due to the fact that the MESE
uses a model based estimation procedure.


\subsection{The Maximum Entropy Spectral Estimate}

	The maximum entropy spectral estimate (MESE) is designed
to produce high-resolution, low-bias spectral estimates
from finite length discrete signals.
The formulation of the MESE problem is as follows.
It is assumed that only a finite number, $N$, of autocorrelation
lags (estimated from a finite length discrete signal) are
available.  The MESE yields the function $\hat{S}_x(\omega)$
which has maximum entropy and whose inverse Fourier transform exactly matches the $N$ lags,
$\hat{r}_x(n)$.  This can be expressed by the equation
%
\begin{equation}
     \hat{S}_x(\omega) = \max_{S(\omega)}\{-\int_{-\pi}^{\pi}S(\omega)\log[S(\omega)]d\omega\}
\label{e3.5}
\end{equation}
%
where
%
\begin{equation}
\begin{array}{cc}
{\displaystyle \frac{1}{2\pi}\int_{-\pi}^{\pi}\hat{S}_x(\omega)e^{j\omega n}d\omega = \hat{r}_x(n)}, & \mbox{$n=0,1,\ldots,N-1$}.
\label{e3.6}
\end{array}
\end{equation}
%
Equation (\ref{e3.5}) expresses the optimality condition of maximizing the
entropy of $S(\omega)$ subject to the $N$ constraints posed in (\ref{e3.6}).

	Since entropy is a measure of randomness, the
MESE is the spectral estimate which is maximally random
given the constraints in (\ref{e3.6}).  Intuitively, the MESE incorporates
no information in the estimated spectrum other than the knowledge
of the autocorrelation lags.  That is to say that the bias
should be eliminated (or at least minimized in some sense)
since no non-data related constraints are imposed on the spectrum.
As was discussed in the introduction, windowed-autocorrelation
spectral estimates suffered from bias due
to the leakage from the window sidelobes.
The window imposes a non-data related constraint on the
power spectrum estimate in that the autocorrelation function
is assumed to be identically zero outside of the support of 
the window.

	Furthermore, as is discussed in \cite{KM},
it can be shown that the MESE is equivalent to the 
Fourier transform of an infinite length autocorrelation
sequence which is obtained by extrapolating the 
sequence of length $N$ in (\ref{e3.6}).  The extrapolation is accomplished
using an auto-regressive, all-pole model of order $N-1$ given
by
%
\begin{equation}
\begin{array}{cc}
{\displaystyle \hat{r}_x(n) = -\sum_{k=1}^{N-1}a_k\hat{r}_{n-k}}, & n\ge N.
\label{e3.7}
\end{array}
\end{equation}
%
Any autocorrelation sequence can be modeled by (\ref{e3.7}) given a large enough model order, $N$.
Consequently, in principle, the resolution of the MESE should
be much better than that of a windowed
spectral estimate since the MESE uses an infinite length autocorrelation sequence.

	The solution to (\ref{e3.5}) and (\ref{e3.6}) can be found
by using the calculus of variations \cite{HB} and, as demonstrated in \cite{KM},
the solution takes the form
%
\begin{equation}
     \hat{S}_x(\omega)=
\frac{\sigma^2}{|1+\sum_{n=1}^{N-1}a_n\exp\{-j\omega n\}|^2}
\label{e3.8}
\end{equation}
%
where the parameter set $\{\sigma^2, a_1, a_2, \ldots, a_{N-1}\}$
is obtained by solving the system of linear equations
%
\begin{equation}
\left[ \begin{array}{cccc}
\hat{r}_x(0) & \hat{r}_x(1) & \cdots & \hat{r}_x(N-1) \\
\hat{r}_x(1) & \hat{r}_x(0) & \cdots & \hat{r}_x(N-2) \\
\vdots       & \vdots       &        & \vdots \\
\hat{r}_x(N-1) & \hat{r}_x(N-2) & \cdots & \hat{r}_x(0) 
\end{array}\right]
\left[ \begin{array}{c}
1\\
a_1\\
\vdots\\
a_{N-1}
\end{array}\right]
=
\left[ \begin{array}{c}
\sigma^2\\
0\\
\vdots\\
0
\end{array}\right]
\label{e3.9}
\end{equation}
%
where the Toeplitz matrix in (\ref{e3.9}) is composed of the $N$ estimated
correlation lags $\hat{r}_x(n)$.  The system of $N$ linear equations in (\ref{e3.9})
are known as the Yule-Walker equations and an
efficient algorithm for their solution is described in the next section.

\subsection{The Levinson Algorithm}
\label{levalgo}
\index{Levinson's algorithm}

	An efficient recursive solution to the Yule-Walker
equations in (\ref{e3.9}) exists and is known as the Levinson
algorithm.  The algorithm requires $O(N^2)$
complex multiplications and additions.  The solution to the 
$k^{th}$ order problem is obtained from the
solution to the $(k-1)^{th}$ order problem using the following
equations
%
\begin{eqnarray}
a_{kk} &=& -[\hat{r}_x(k)+\sum_{j=1}^{k-1}a_{k-1,j}\hat{r}_x(k-j)]/\sigma^2_{k-1}\\
a_{ki} &=& a_{k-1,i}+a_{kk}a^*_{k-1,k-i}\\
\sigma^2_k &=& (1-|a_{kk}|^2)\sigma^2_{k-1}.
\end{eqnarray}
%
The solution to the $1^{st}$ order problem is
%
\begin{eqnarray}
a_{11} &=& -\hat{r}_x(1)/\hat{r}_x(0)\\
\sigma^2_1 &=& (1-|a_{11}|^2)\hat{r}_x(0).
\end{eqnarray}
%
\subsection{How to Use {\tt mese}}
\index{macro syntax!mese@{\tt mese}}
\index{maximum entropy method!macro syntax}

	The syntax for the macro {\tt mese} is as follows,
\begin{verbatim}
-->[sm,fr]=mese(x)
\end{verbatim}
where one wants to obtain a power spectral estimate of
{\tt x}, the input data sequence, and {\tt sm} is the resulting
estimate obtained on the normalized frequency axis ($0\le${\tt fr}$\le .5$).

\subsection{How to Use {\tt lev}}
\index{macro syntax!lev@{\tt lev}}
\index{Levinson's algorithm!macro syntax}

	The syntax for the macro {\tt lev} is as follows,
\begin{verbatim}
-->[ar,sigma2,rc]=lev(r)
\end{verbatim}
where {\tt r} is a vector of auto-correlation coefficients ($r(0),r(1),
\ldots,r(N-1)$), {\tt ar} is the vector which satisfies the Yule-Walker
equations, {\tt sigma2} is the scalar which satisfies the Yule-Walker
equations, and {\tt rc} is a vector of reflection coefficients.

\subsection{Examples}
\index{maximum entropy method!examples}

	Here we give an example of estimating the power spectrum of 
a very short data sequence using the MESE and also using the magnitude
squared of the Fourier transform.  The data is eleven
samples of the sum of two 
sinusoids in additive, uniformly distributed, white noise.  The functional
form of the data sequence is
%
\begin{equation}
x(n)=\sin(2\pi n/20)+\sin(3.5\pi n/20)+.2w(n)
\end{equation}
%
where $w(n)$ is the white noise sequence which takes on values in $[-1,1]$
and $n=0,1,...,10$.  Figure~\ref{f33.1} shows the input data sequence, x(n).
Figures~\ref{f3.2} and \ref{f3.3} show the maximum entropy and magnitude
squared estimates of the power spectrum, respectively.
%
\input{\Figdir mem1.tex}
\dessin{{\tt exec('mem1\_3.code')} Input Data Sequence, $x(n)$}{f33.1}
%
%
\input{\Figdir mem2.tex}
\dessin{{\tt exec('mem1\_3.code')} Maximum Entropy Spectral Estimate of $x(n)$}{f3.2}
%
\input{\Figdir mem3.tex}
\dessin{{\tt exec('mem1\_3.code')} Squared Magnitude of the Fourier Transform of $x(n)$}{f3.3}
%
As can be seen, the MESE resolves two peaks according to the 
two sinusoidal frequences in $x(n)$.  The squared magnitude of
the Fourier transform of $x(n)$ does not have a long enough signal
to resolve the two peaks of the spectrum.  Furthermore, the
power spectrum estimate in Figure~\ref{f3.3} shows spurious sidelobes
which have nothing to do with the data.



%\end{document}
