% Copyright INRIA

\chapter{Introduction}

\section{What is Scilab}

Developed at INRIA, Scilab has been developed for system control and
signal processing applications. It 
is freely distributed in source code format (see the copyright file). 

Scilab is made of three distinct parts: an interpreter, 
libraries of functions (Scilab procedures) and libraries of Fortran 
and C routines. 
These routines (which, strictly speaking, do not belong to Scilab but
are interactively called by the interpreter) are of 
independent interest and most of them are available through Netlib. 
A few of them have been slightly modified for better compatibility
with Scilab's interpreter.
 
A key feature of the Scilab syntax is its ability to handle matrices: 
basic matrix manipulations such as concatenation, 
extraction or transpose are immediately performed as well as basic operations
such as addition or multiplication. Scilab also aims at handling more complex
objects than numerical matrices. For instance, control people may want
to manipulate rational or polynomial transfer matrices. This is
done in Scilab by manipulating lists and typed lists which allows a 
natural symbolic representation of complicated
mathematical objects such as transfer functions, linear systems or graphs
(see Section~\ref{s2.5}).

        Polynomials, polynomials matrices and transfer matrices 
are also defined and the syntax used for manipulating these matrices
is identical to that used for manipulating constant vectors and matrices.

        Scilab provides a variety of powerful primitives for
the analysis of non-linear systems.  
Integration of explicit and implicit dynamic systems can be accomplished 
numerically.  The {\tt scicos} toolbox allows the graphic definition
and simulation of complex interconnected hybrid systems.

 There exist numerical optimization 
facilities for non linear optimization (including
non differentiable optimization), quadratic optimization and 
linear optimization.

        Scilab has an open programming environment where the
creation of functions and libraries of functions is completely in the
hands of the user (see Chapter~\ref{ch4}).    
Functions are recognized as data objects in Scilab and, thus, can be 
manipulated or created as other data objects.  For example, functions
can be defined inside Scilab and passed as input or output 
arguments of other functions.

In addition Scilab supports a character string data type 
which, in particular, allows the on-line creation of functions.
Matrices of character strings are also manipulated with the same
syntax as ordinary matrices.

        Finally, Scilab is easily interfaced with Fortran or C 
subprograms.  This allows use of standardized 
packages and libraries in the interpreted environment of Scilab.

        The general philosophy of Scilab is to provide the following
sort of computing environment:
\begin{itemize}
   \item To have data types which are varied and flexible with 
a syntax which is natural and easy to use.
   \item To provide a reasonable set of primitives which serve
           as a basis for a wide variety of calculations.
   \item To have an open programming environment where new
           primitives are easily added. A useful tool distributed with
 Scilab is {\tt intersci} which
is a tool for building interface programs to add new primitives 
i.e. to add new modules of Fortran or C code into Scilab.

   \item To support library development through ``toolboxes'' of
         functions devoted to specific
           applications (linear control, signal processing, 
           network analysis, non-linear control, etc.)
\end{itemize}

        The objective of this introduction manual is to give the user 
an idea of what Scilab can do. On line documentation on all
functions is available ({\tt help} command).


\section{Software Organization}

Scilab is divided into a set of directories. The main directory
\verb!SCIDIR! contains the following files: {\tt scilab.star} (startup file), the
copyright file {\tt notice.tex}, and the \verb!configure! files
(see(\ref{install})).
The subdirectories are the following:
\begin{itemize}

\item{{\tt bin} is the directory of the executable files.
The starting script {\tt scilab} on Unix/Linux systems and {\tt runscilab.exe} on Windows95/NT, The executable code of Scilab: {\tt scilex} on Unix/Linux systems and {\tt scilex.exe} on Windows95/NT are there. 
This directory also contains Shell scripts 
for managing or printing Postscript/\LaTeX\  files produced by Scilab.}

\item{{\tt demos} is the directory of demos. 
This directory contains the codes corresponding
to various demos. They are often useful for inspiring new users.
The file 
{\tt alldems.dem} is used by the  ``Demos'' button. 
Most of plot commands are illustrated by simple demo examples. 
Note that running a graphic function without input parameter
provides an example of use for this function (for instance {\tt
plot2d()} displays an example for using {\tt plot2d}  function). }

\item{{\tt examples} contains useful examples of how to link
external programs to scilab, using dynamic link or {\tt intersci}}

\item{{\tt doc} is the directory of the Scilab documentation: \LaTeX\ , dvi 
and Postscript files. 
This documentation is {\tt SCIDIR/doc/intro/intro.tex}. }

\item{{\tt geci} contains source code and binaries for GeCI which is an 
interactive communication manager created in order to
manage remote executions of softwares and allow exchanges of messages
beetwen those programs. It offers the possibility to exploit numerous
machines on a network, as a virtual computer, by creating a
distributed group of independent softwares 
(\verb!help communications! for a detailed description).
GeCI is used for the link of Xmetanet with Scilab.}

\item{\tt pvm3} contains source code and binaries of the PVM version
3 which is another interactive communication manager. 

\item{{\tt imp} is the directory of the routines managing the Postscript files
for print.}

\item{{\tt libs} contains the Scilab libraries (compiled code).}

\item{{\tt macros} contains the libraries of functions
which are available on-line. New libraries can easily be added 
(see the Makefile). This directory is divided into a number of subdirectories
which contain ``Toolboxes'' for control, signal processing, etc... Strictly
speaking Scilab is not organized in toolboxes : functions of a specific
subdirectory can call functions of other directories; so, for example, the 
subdirectory {\tt signal} is not self-contained but its functions are all devoted
to signal processing.}

\item{{\tt man} is the directory containing the manual divided 
into submanuals, corresponding to the on-line help and to 
a \LaTeX\  format of the reference manual. 
The \LaTeX\  code is produced by a 
translation of the Unix format Scilab manual (see the subdirectory
{\tt SCIDIR/man}).}
To get information about an item, one should enter 
{\tt help item}\index{help@{\tt help}} 
in Scilab or use the help window facility obtained with help button.
To get information corresponding to a key-word, one should  enter 
{\tt apropos key-word} or use 
{\tt apropos}\index{apropos@{\tt apropos}} in the help window. 
All the {\tt item}s and {\tt key-words} known by the {\tt help} and 
{\tt apropos} commands are in {\tt .cat} and {\tt whatis} files 
located in the {\tt man} subdirectories.

To add new items to the {\verb!help!} and {\tt apropos} commands 
the user can extend the list of directories available to the help 
browser by adapting the variable \verb!%helps!. 
See the README file in the \verb!man! directory and the example
given in \verb!examples/man-examples! directory

\item{{\tt maple} is the directory which contains the source code of Maple
functions which allow the transfer of Maple objects
into Scilab functions. For efficiency, the transfer
is made through Fortran code generation which is dynamically linked
to Scilab.}

\item{{\tt routines} is a directory which contains the source code of all
the numerical routines. The subdirectory {\tt default} is important
since it contains the
source code of routines which are necessary to customize Scilab.}
In particular user's C or Fortran routines for ODE/DAE simulation 
or optimization can be included here (they can be also dynamically
linked).

\item{{\tt examples} contains examples of specific topics. It is shown
in appropriate subdirectories how to add new C or Fortran program
to Scilab (see \verb!addinter-tutorial!). More complex examples are given
in \verb!addinter-examples!. The directory \verb!mex-examples!
contains examples of interfaces realized by emulating the Matlab
mexfiles. The directory \verb!link-examples! illustrates the use of
the \verb!call! function which allows to call external function within
Scilab.}

\item{{\tt intersci} contains a program which can be used to build
interface programs for adding new Fortran or C 
primitives to Scilab. This program is executed by the {\tt intersci}
script in the {\tt bin/intersci} directory.}

\item{{\tt scripts} is the directory which contains the source code of 
shell scripts files. Note that the list of printers \index{printers} 
names known by Scilab is defined there by an environment variable.}

\item{{\tt tests} : this directory contains evaluation programs for testing 
Scilab's installation on a machine. The file ``demos.tst'' tests all the 
demos.}

\item{{\tt wless, xless} is the Berkeley file browsing tool}

\item{{\tt xmetanet} is the directory which contains {\tt xmetanet}, a 
graphic display for networks. Type {\tt metanet()} in Scilab to use it.}

\end{itemize}

\section{Installing Scilab. System Requirements}
\label{install}
Scilab is distributed in source code format; binaries for Windows95/NT
systems and several popular Unix/Linux-XWindow systems are also
available: Dec Alpha (OSF V4), Dec Mips (ULTRIX 4.2), Sun Sparc
stations (Sun OS), Sun Sparc stations (Sun Solaris), HP9000 (HP-UX
V10), SGI Mips Irix, PC Linux. All of these binaries versions include
tk/tcl interface.

The installation requirements are the following :

- for the source version: Scilab requires approximately 130Mb of disk
storage to unpack and install (all sources included).  You need X
Window (X11R4, X11R5 or X11R6, C compiler and Fortran compiler
(e.g. f2c or g77 or Visual C++ for Windows systems). 
  
- for the binary version: the minimum for running Scilab (without
sources) is about 40 Mb when decompressed.
These versions  are partially statically linked and in principle do not require a fortran compiler.

\label{stks}
Scilab uses a large internal stack for its calculations. This size
of this stack can be reduced or enlarged by the 
{\tt stacksize}\index{stacksize@{\tt stacksize}}.
command. The default dimension of the internal stack can be adapted
by modifying the variable {\tt newstacksize} in the {\tt scilab.star}
script.

- For more information on the installation, please look at the README files

\section{Documentation}

The documentation is made of this User's guide (Introduction to
Scilab) and the Scilab on-line manual. There are also
reports devoted to specific toolboxes: Scicos (graphic system builder
and simulator), Signal (Signal processing toolbox), Lmitool (interface for 
LMI problems), Metanet (graph and network toolbox). An FAQ is
available at Scilab home page:\\
(\verb!http://www-rocq.inria.fr/scilab!)\index{home page}.

\section{Scilab at a Glance. A Tutorial}

\subsection{Getting Started}

Scilab is called by running the {\tt scilab} script in the 
directory {\tt SCIDIR/bin} ({\tt SCIDIR} denotes the directory 
where Scilab is installed).
This shell script  runs Scilab in an Xwindow environment (this 
script file can be invoked with
specific parameters such as \verb!-nw! for ``no-window'').
You will immediatly get the Scilab window with the following banner and 
prompt represented by the {\tt -->} : 


\bigskip

\begin{verbatim}
                           ===========
                           S c i l a b
                           ===========
 
 
                  Scilab-2.x  ( 12 July 1998 ) 
                  Copyright (C) 1989-98 INRIA 
 
 
Startup execution:
  loading initial environment
   
 -->

\end{verbatim}


A first contact with Scilab can be made by clicking 
on {\tt Demos} with the left mouse button and clicking then on 
{\tt Introduction to SCILAB }: the 
execution of the session is then done by entering empty lines and can be
stopped with the buttons {\tt Stop} and {\tt Abort}.

  Several libraries
(see the {\tt SCIDIR/scilab.star} file) are automatically loaded.
        
To give the user an idea of some of the capabilities of Scilab
we will give later a sample session in Scilab.\\

\subsection{Editing a command line}

Before the sample session, we briefly present how to edit a command line.
You can enter a command line by typing after the prompt or clicking with the 
mouse on a part on a window and copy it at the prompt in the Scilab
window. 
The pointer may be moved using the directionnal arrows
($\leftarrow^\uparrow_\downarrow \rightarrow$). For Emacs customers,
the usual Emacs commands are at your 
disposal for modifying a command (Ctrl-$<$chr$>$  means hold the CONTROL key 
while typing the character $<$chr$>$), for example:

\bigskip


%
\begin{itemize}
\item Ctrl-p    recall previous line
\item Ctrl-n    recall next line
\item Ctrl-b    move backward one character
\item Ctrl-f    move forward one character
\item Delete    delete previous character
\item Ctrl-h    delete previous character
\item Ctrl-d    delete one character (at cursor)
\item Ctrl-a    move to beginning of line
\item Ctrl-e    move to end of line
\item Ctrl-k    delete to the end of the line
\item Ctrl-u    cancel current line
\item Ctrl-y    yank the text previously deleted
\item !{\tt prev}     recall the last command line which begins by {\tt prev}
\item Ctrl-c    interrupt Scilab and pause after carriage return. 
Clicking on the Control/stop button enters a Ctrl-c.
\end{itemize}
%

As said before you can also cut and paste using the mouse. This way will be
useful if you type your commands in an editor. Another way to 
``load'' files containing Scilab statements
is available with the {\tt File/File Operations} button.

\subsection{Buttons}
The Scilab window has the following {\bf Control} buttons.\\

%
\begin{itemize}
\item Stop     interrupts execution of Scilab and enters in 
{\tt pause} mode
\item Resume    continues execution after a {\tt pause}
entered as a command in a function or generated by the {\tt Stop}
button or Control C.
\item Abort     aborts execution after one (or several) {\tt pause}, 
and returns to top-level prompt
\item Restart   clears all variables and executes startup files
\item Quit      quits Scilab
\item Kill      kills Scilab shell script
\item Demos     for interactive run of some demos
\item File Operations   facility for loading functions or data into
Scilab, or executing script files. 
\item Help : invokes on-line help with the tree of the man and the 
names of the corresponding items. It is possible to type directly 
{\tt help <item>} in the Scilab window.
\item Graphic Window : select active graphic window
\end{itemize}
%
New buttons can be added by the 
{\tt addmenu}\index{addmenu@{\tt addmenu}} command.
Note that the command:\\
{\tt SCIDIR/bin/scilab -nw} \\
invokes Scilab in the ``no-window'' mode.

\subsection{Customizing your Scilab - Unix only}


The parameters of the different windows opened by 
Scilab can be easily changed. The way for doing that is to edit the files 
contained in the directory {\tt X11-defaults}. The first possibility is 
to directly customize these files. Another way is to copy 
the right lines with the modifications in the {\tt .Xdefaults} 
file of the home directory.
These modifications are activated by starting again Xwindow or with the
command {\tt xrdb .Xdefaults}. Scilab will read the {\tt .Xdefaults} file: 
the lines of this
file will cancel and replace the corresponding lines of X11-defaults.

A simple example : 

\begin{verbatim}
Xscilab.color*Scrollbar.background:red
Xscilab*vpane.height: 500
Xscilab*vpane.width:  500
\end{verbatim}

in {\tt .Xdefaults} will change the 500x650 window to a square window of 
500x500 and the  
scrollbar background color changes from green to red.

An important parameter for customizing Scilab is \verb!stacksize!
discussed in \ref{install}.

\subsection{Sample Session for Beginners}

We present now some simple commands. At the carriage return all the 
commands typed since the last prompt are interpreted. 

\noindent\dotfill

\input{diary/d1p1.dia}

Give the values of 1 and 2 to the variables {\tt a} and { \tt A} .  The 
semi-colon at the end of the command suppresses the display of the result.
Note that Scilab is case-sensitive.
Then two commands are processed and the second result is displayed because
it is not followed by a semi-colon. The last command shows how to write a
command on several lines by using ``{\tt ...}''. This sign is only needed
in the on-line typing for avoiding the effect of the carriage return.
 The chain of characters which follow the {\tt //} is not interpreted 
(it is a comment line).

\noindent\dotfill

\input{diary/d1p2.dia}

We get the list of previously defined variables {\tt a b c A}  together
with the initial environment composed of the different libraries and
some specific ``permanent'' variables.

Below is an example of an expression which mixes constants with existing
variables.  The result is retained in the standard default variable 
{\tt ans}\index{ans@{\tt ans}}.

\noindent\dotfill

\input{diary/d1p2bis.dia}

Defining {\tt I}, a vector of indices, {\tt W} a random 2 x 4 matrix,
and extracting submatrices from {\tt W}. The \verb!$! symbol stands
for the last row or last column index of a matrix or vector. The colon
symbol stands for ``all rows'' or ``all columns''.

\noindent\dotfill

\input{diary/d1p3.dia}

Calling a function (or primitive) with a vector argument.  The response
is a complex vector.

\noindent\dotfill

\input{diary/d1p4.dia}

A  more complicated command which creates a polynomial. 

\noindent\dotfill

\input{diary/d1p5.dia}

Definition of a polynomial matrix. The syntax for polynomial matrices
is the same as for constant matrices. Calculation of the
determinant of the polynomial matrix by the {\tt det} function. 


\noindent\dotfill

\input{diary/d1p6.dia}

Definition of a matrix of rational polynomials. (The internal representation
of {\tt F} is a typed list of the form {\tt tlist('the type',num,den)}
where {\tt num} and {\tt den} are two matrix polynomials). Retrieving
the numerator and denominator matrices of {\tt F} by extraction operations in a
typed list. Last command is the direct extraction of entry {\tt 1,2} 
of the numerator matrix {\tt F.num}.

\noindent\dotfill

\input{diary/d1p7.dia}


Here we move into a new environment using the command 
{\tt pause}\index{pause@{\tt pause}}
and we obtain the new prompt {\tt -1->} which indicates the level
of the new environment (level 1).  All variables that are available
in the first environment are also available in the new environment.  Variables
created in the new environment can be returned to the original environment
by using {\tt return}\index{return@{\tt return}}.  
Use of {\tt return} without an argument 
destroys all the variables created in the new environment before returning
to the old environment. The {\tt pause} facility is very useful 
for debugging purposes.

\noindent\dotfill

\input{diary/d1p8.dia}

Definition of a rational polynomial by extraction of an entry
of the matrix {\tt F} defined above.  This is followed by the evaluation
of the rational polynomial at the vector of complex frequency values defined
by {\tt frequencies}.  The evaluation of the rational polynomial is done by
the primitive {\tt freq}. {\tt F12.num} is the numerator
polynomial and {\tt F12.den} is the denominator polynomial of the
rational polynomial {\tt F12}. Note that
the polynomial {\tt F12.num} can be also obtained by extraction
from the matrix {\tt F} using the syntax {\tt F.num(1,2)}.
The  visualization of the resulting evaluation
is made by using the basic plot command {\tt plot2d} (see Figure~\ref{f1.1}).

\noindent\dotfill

\input{diary/d1p9.dia}

The function {\tt horner} performs a (possibly symbolic) change of 
variables for a polynomial (for example, here, to
perform the bilinear transformation f(w(s))).

\noindent\dotfill

\input{diary/d1p10.dia}

Definition of a linear system in state-space representation.
The function {\tt syslin} defines here the continuous time ({\tt 'c'}) system
{\tt Sl} with state-space matrices ({\tt A,B,C}). The function
{\tt ss2tf} transforms {\tt Sl} into transfer matrix representation.

\noindent\dotfill

\input{diary/d1p11.dia}

Definition of the rational matrix {\tt R}. {\tt Sl} is the
continuous-time linear system with (improper) transfer matrix
{\tt R}. {\tt tf2ss} puts {\tt Sl} in state-space representation with a
polynomial {\tt D} matrix. Note that linear systems are represented
by specific typed lists (with 7 entries).

\noindent\dotfill

\input{diary/d1p12.dia}

{\tt sl1} is the linear system in transfer matrix representation
obtained by the parallel inter-connection of {\tt Sl} and {\tt 2*Sl +eye()}.
The same syntax is valid with {\tt Sl} in state-space representation.

\noindent\dotfill

\input{diary/d1p13.dia}

On-line definition of a function, called {\tt compen} which calculates the 
state space representation
({\tt Cl}) of a linear system ({\tt Sl}) controlled by an observer
with gain {\tt Ko}
and a controller with gain {\tt Kr}.  Note that matrices are constructed
in block form using other matrices.

\noindent\dotfill

\input{diary/d1p14.dia}

Call to the function {\tt compen} defined above where the gains were
calculated by a call to the primitive {\tt ppol} which performs pole
placement.
The resulting {\tt Aclosed} matrix is displayed and the placement
of its poles is checked using the primitive {\tt spec} which calculates
the eigenvalues of a matrix. (The function {\tt compen} is defined here
on-line by  as an example of function which receive a linear system 
({\tt Sl}) as input and returns a linear system ({\tt Cl}) as output.
In general Scilab functions are defined in files and loaded in Scilab
by {\tt exec} or by {\tt getf} ).

\noindent\dotfill

\input{diary/d1p15.dia}

Relation with the Unix environment.

\noindent\dotfill

\input{diary/d1p16.dia}

Definition of a column vector of character strings used for defining a C
function file. The routine is compiled (needs a compiler), dynamically 
linked to Scilab by the {\tt link} command, and interactively called
by the function {\tt myplus}.

\noindent\dotfill

\input{diary/d1p17.dia}

Definition of a function which calculates a first order vector differential
{\tt f(t,y)}.  This is followed by the definition of the constant {\tt a}
used in the function.  The primitive {\tt ode}
then integrates the differential equation defined by the Scilab
function {\tt f(t,y)}
for {\tt y0=[1;0]} at {\tt t=0} and where the solution is given
at the time values $t=0,.02,.04,\ldots,20$. (Function {\tt f} can
be defined as a C or Fortran program). The result is plotted in
Figure~\ref{f1.2} where the first element of the integrated vector is 
plotted against the second element of this vector.

\noindent\dotfill

\input{diary/d1p18.dia}

Definition of a matrix containing character strings. By default, the 
operation of symbolic multiplication of two matrices of character
strings is not defined in Scilab.  However, the (on-line)
function definition for {\tt \%cmc} defines the multiplication of 
matrices of character strings.
The {\tt \%} which begins the function definition for {\tt \%cmc}
allows the definition of an operation which did not previously 
exist in Scilab, and the name {\tt cmc} means ``chain multiply chain''.
This example is not very useful: it is simply given to show
how {\em operations} such as \verb!*! can be defined on complex data 
structures by mean of scpecific Scilab functions.

\noindent\dotfill

\input{diary/d1p19.dia}

A simple example which illustrates the passing of a function as an argument
to another function. Scilab functions are objects which may be defined, loaded,
or manipulated as other objects such as matrices or lists.

\noindent\dotfill

\begin{verbatim}
-->quit
\end{verbatim}

Exit from Scilab.

\noindent\dotfill

\input{figures/d1-7.tex}
\caption{\label{f1.1}A Simple Response}
\end{figure}

\input{figures/d1-14.tex}
\caption{\label{f1.2}Phase Plot}
\end{figure}

